Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-05-10 src/tex/ms.tex
replace the term "super-"runaway by runaway

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
With even a fraction of the Earth's water inventory, a super-runaway planet will be in a global magma ocean state~\citep{Lichtenberg2021c,Boukrouche2021}.
We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,Salvador2023}.

@@

For \textit{super-runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.

@@

We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population.
(i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
(iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
It has been suggested that the atmospheric depth and the presence or absence of oceans on ``sub-Neptunes'' could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
This has been suggested to be done via (v) eclipse photometry~\citep{Mansfield2019,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.

@@

Other missions have been proposed that focus on characterizing exoplanet habitability, most notably the \hwo\ concept, which will build on the two precursor direct imaging concepts LUVOIR~\citep{LUVOIR2019} and HabEx~\citep{Gaudi2020c}.
Similar science objectives are pursued by the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2022} initiative, a mission concept utilizing a space-based mid-infrared nulling interferometer.
Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.
However, because mid-infrared retrievals feature reduced degeneracy between cloud albedo and changes in surface area, the planet radius can be constrained in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
Since the peak thermal emission in super-runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of super-runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Sect.~\ref{sec:results_FGK_M}).
With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states~\citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.

@@

Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G}.

>>>>

\section{Runaway Greenhouse Model}\label{sec:met_rghmodel}
%\todo[inline]{
%\textbf{Main assumptions:}\\
%- every planet (modulo $f_\mathrm{rgh}$) has \textit{some} water. Enough for runaway GH under the right conditions.  \\
%- Limitation to small planets without significant atmospheric envelopes.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets within the runaway greenhouse regime~\citep{Lichtenberg2022}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%}
The climate state of a planet has a direct influence on its apparent size measured by transit photometry~\citep{Turbet2019,Turbet2020,Mousis2020,2021ApJ...914...84A}.
With even a fraction of the Earth's water inventory, a runaway planet will be in a global magma ocean state~\citep{Lichtenberg2021c,Boukrouche2021}.
We use predictions on transit atmospheric thickness from geophysical models to derive the change in transit radius and bulk density that planets with instellation-induced runaway greenhouse climates experience, depending on the thermal stratification of the atmosphere and dissolution of water into the underlying magma ocean~\citep{Dorn2021,Salvador2023}.

@@

For \textit{runaway} planets, i.e., all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.

@@

We currently see four potential lines of discriminating atmospheric signatures associated with runaway climates in the exoplanet population.
(i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
In a runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this runaway regime~\citep{2020SSRv..216...98G}.
(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
(iv) Disequilibrium chemistry in tidally-locked runaway planets.
It has been suggested that the atmospheric depth and the presence or absence of oceans on ``sub-Neptunes'' could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
This has been suggested to be done via (v) eclipse photometry~\citep{Mansfield2019,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.

@@

Other missions have been proposed that focus on characterizing exoplanet habitability, most notably the \hwo\ concept, which will build on the two precursor direct imaging concepts LUVOIR~\citep{LUVOIR2019} and HabEx~\citep{Gaudi2020c}.
Similar science objectives are pursued by the Large Interferometer For Exoplanets~\citep[\life,][]{Quanz2022} initiative, a mission concept utilizing a space-based mid-infrared nulling interferometer.
Direct imaging surveys do not directly measure planetary radii and are primarily useful for providing context through atmospheric measurements of individual planets.
However, because mid-infrared retrievals feature reduced degeneracy between cloud albedo and changes in surface area, the planet radius can be constrained in mid-infrared wavelengths~\citep{2018ExA....46..543D,2021ExA...tmp..118Q}.
Mid-infrared direct imaging techniques, in particular, enable to study much deeper atmospheric layers than possible in reflected light~\citep{Wordsworth2022}.
Hence, the atmospheric structure can be retrieved for a wider variety of thermal and atmospheric scenarios~\citep{Alei2022,Konrad2022}.
Since the peak thermal emission in runaway climates will substantially decrease the star-to-planet flux ratio, mid-infrared wavelengths offer the possibility to probe the diversity of runaway climates in systems across different ages~\citep{2014ApJ...784...27L,2019A&A...621A.125B}.
Finally, mid-infrared surveys such as \life\ show a preference for M~star planets~\citep{Quanz2022}, which is beneficial for detecting the predicted habitable zone inner edge discontinuity (see Sect.~\ref{sec:results_FGK_M}).
With a sample size of a few tens of planets crossing the runaway greenhouse transition, direct imaging missions will thus enable key insights into the compositional inventory of atmospheric volatiles and climate states~\citep{2021exbi.book....5H,2022arXiv220505696C}, adding important details to a potential runaway greenhouse detection purely via transit radii.

@@

Future self-consistent modeling of interior-atmosphere interactions may include constraints on additional radius increases due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,Lichtenberg2021c,2022PSJ.....3...93B,2021SSRv..217...22G}.
